---
title: "CDC Diabetes Health Indicators Analysis"
author: "Josie Charles, Ray Prasad, Dawson Ren"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(glmnet)
library(caret)
library(corrplot)
library(pROC)
library(rpart)
library(rpart.plot)
```

# Introduction

This project analyzes the CDC Diabetes Health Indicators dataset to predict whether an individual has diabetes based on various lifestyle factors and health indicators. The goal is to build a binary classification model that can accurately identify individuals at risk of diabetes. We will explore the data, handle multicollinearity, and compare different logistic regression models, including those with regularization (LASSO and Ridge).

# Data Description

The dataset is obtained from the UCI Machine Learning Repository (ID: 891). It contains health indicators for 253,680 individuals.

## Raw Data and Pre-processing

The raw data was fetched using the `ucimlrepo` Python package and exported to a CSV file. The dataset includes a binary target variable `Diabetes_binary` (0 = no diabetes, 1 = pre-diabetes or diabetes) and 21 feature variables.

```{r load_data}
# Load data
data <- read.csv("diabetes_data.csv")

# Convert target to factor
data$Diabetes_binary <- as.factor(data$Diabetes_binary)
```

The dataset contains `r nrow(data)` observations and `r ncol(data)` variables. There are no missing values in the dataset.

# Exploratory Data Analysis (EDA)

Let's look at the distribution of the target variable.

```{r eda_target, fig.height=4, fig.width=6}
ggplot(data, aes(x = Diabetes_binary)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Diabetes Status", x = "Diabetes (0 = No, 1 = Yes)", y = "Count") +
  theme_minimal()
```

Notice that there is significant class imbalance, only `r mean(data$Diabetes_binary)` percent of interviewed people had diabetes. We will look at up/downweighting to deal with the class imbalance.

We can also examine the correlation between numeric predictors to identify potential multicollinearity.

```{r correlation, fig.height=8, fig.width=8}
# Calculate correlation matrix for numeric predictors
numeric_vars <- data %>% select(-Diabetes_binary)
cor_matrix <- cor(numeric_vars)
corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black", tl.cex = 0.7)
```

```{r correlation_pairs, fig.height=8, fig.width=8}
# Find and print pairs with specific correlation strengths
cor_melted <- as.data.frame(as.table(cor_matrix))
names(cor_melted) <- c("Var1", "Var2", "Correlation")

# Filter out self-correlations and duplicate pairs
cor_melted <- cor_melted %>%
  filter(as.character(Var1) < as.character(Var2))

print_correlation_pairs <- function(df, lower_bound, upper_bound, strength_label) {
  pairs <- df %>%
    filter(abs(Correlation) >= lower_bound & abs(Correlation) < upper_bound) %>%
    arrange(desc(abs(Correlation)))

  if (nrow(pairs) > 0) {
    cat(paste0("\n--- ", strength_label, " Correlations (", lower_bound, "-", upper_bound, ") ---\n"))
    for (i in 1:nrow(pairs)) {
      cat(sprintf("%s - %s: %.2f\n", pairs$Var1[i], pairs$Var2[i], pairs$Correlation[i]))
    }
  } else {
    cat(paste0("\n--- No ", strength_label, " Correlations found (", lower_bound, "-", upper_bound, ") ---\n"))
  }
}

print_correlation_pairs(cor_melted, 0.2, 0.4, "Weak")
print_correlation_pairs(cor_melted, 0.4, 0.6, "Moderate")
print_correlation_pairs(cor_melted, 0.6, 1.0, "Strong")
```

The strongest correlations are between GenHlth/PhysHlth/DiffWalk, and Education/Income. This makes sense,
as people with better general health likely have better physical health and less difficulty walking.
And people with higher education are likely to have a higher income.

We also calculate the variance inflation factor (VIF) to better identify multicollinearity.

```{r vif, fig.height=8, fig.width=8}
# Calculate by inverting the correlation matrix
X <- data %>% select(-Diabetes_binary)
# turn X into a matrix
X <- as.matrix(X)
# calculate the correlation matrix
cor_matrix <- cor(X)
# calculate the VIF
vif <- diag(solve(cor_matrix))

# Visualize VIF
barplot(vif, main = "Variance Inflation Factor", xlab = "Variable", ylab = "VIF", horiz = TRUE)
```

It seems that the VIFs are relatively small, so we do not need to remove any variables.

# Methodology

We will use K-fold Cross-Validation (5-fold) for model training and evaluation to ensure stability.
We will also address class imbalance using upsampling and downsampling techniques.
Furthermore, we will use LASSO to select the best variables, including interaction and quadratic terms.

```{r split_data}
set.seed(123)
# Ensure target is a factor with valid names for caret
data$Diabetes_binary <- factor(data$Diabetes_binary, levels = c("0", "1"), labels = c("No", "Yes"))

trainIndex <- createDataPartition(data$Diabetes_binary,
  p = .8,
  list = FALSE,
  times = 1
)
dataTrain <- data[trainIndex, ]
dataTest <- data[-trainIndex, ]
```

## Modeling

We will train three types of models:
1. **Logistic Regression** (Baseline)
2. **Logistic Regression with Class Balancing** (Up/Downsampling)
3. **Logistic Regression with LASSO**

```{r train_control}
# Define train control with 5-fold CV and class probabilities
ctrl <- trainControl(
  method = "cv",
  number = 5,
  summaryFunction = twoClassSummary,
  classProbs = TRUE,
  savePredictions = "final"
)
```

### Logistic Regression

```{r logistic_models}
# Standard Logistic Regression (No sampling)
set.seed(123)
model_glm <- train(Diabetes_binary ~ .,
  data = dataTrain,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  trControl = ctrl
)

# Logistic Regression with Downsampling
ctrl_down <- ctrl
ctrl_down$sampling <- "down"
set.seed(123)
model_glm_down <- train(Diabetes_binary ~ .,
  data = dataTrain,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  trControl = ctrl_down
)

# Logistic Regression with Upsampling
ctrl_up <- ctrl
ctrl_up$sampling <- "up"
set.seed(123)
model_glm_up <- train(Diabetes_binary ~ .,
  data = dataTrain,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  trControl = ctrl_up
)
```

```{r}
# Logistic Regression using LASSO
# We use glmnet for LASSO. We tune lambda. Alpha is set to 1 for LASSO.
set.seed(123)
lasso_grid <- expand.grid(alpha = 1, lambda = 10^seq(-4, -1, length = 100))

vars <- colnames(dataTrain)
vars <- vars[vars != "Diabetes_binary"]
interaction_terms <- paste(vars, collapse = " + ")

# add quadratic terms for interaction terms
formula <- as.formula(
  paste("Diabetes_binary ~ (", interaction_terms, ")^2 + I(BMI^2) + I(MentHlth^2) + I(GenHlth^2) + I(Age^2) + I(Income^2)")
)

model_lasso <- train(formula,
  data = dataTrain,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = ctrl_down, # Using downsampling
  tuneGrid = lasso_grid
)
```

```{r}
# Plot the coefficient path
plot(model_lasso$finalModel, xvar = "lambda", label = TRUE)
title("LASSO Coefficient Path", line = 2.5)
```

```{r}
lasso_coefs <- coef(model_lasso$finalModel, s = model_lasso$bestTune$lambda)
lasso_coefs_vec <- as.vector(lasso_coefs)
names(lasso_coefs_vec) <- rownames(lasso_coefs)
lasso_coefs <- lasso_coefs_vec[names(lasso_coefs_vec) != "(Intercept)"]
lasso_coefs <- sort(lasso_coefs, decreasing = TRUE)
print(lasso_coefs)
```

Select 1SE sparser model.
```{r}
res <- model_lasso$results
best_idx <- which.max(res$ROC)
best <- res[best_idx, ]

roc_se <- best$ROCSD / sqrt(model_lasso$control$number)
threshold <- best$ROC - roc_se

one_se_idx <- max(which(res$ROC >= threshold))
lambda_1se <- res$lambda[one_se_idx]

coef_1se <- coef(model_lasso$finalModel, s = lambda_1se)
coef_1se_vec <- as.vector(coef_1se)
names(coef_1se_vec) <- rownames(coef_1se)
coef_1se <- coef_1se_vec[names(coef_1se_vec) != "(Intercept)"]
coef_1se <- sort(coef_1se, decreasing = TRUE)
print(coef_1se)
```

We only take coefficients with absolute values greater than 0.1. This is nonstandard,
but we do this to identify a very sparse model.
```{r}
coef_1se <- coef_1se[abs(coef_1se) > 0.1]
print(coef_1se)
```

Finally, we refit the model with the selected variables.
```{r}
formula_lasso_1se <- as.formula(
  paste("Diabetes_binary ~ ", paste(names(coef_1se), collapse = " + "))
)
model_lasso_1se <- train(formula_lasso_1se,
  data = dataTrain,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  trControl = ctrl_down
)
``` 

## Residual Analysis

Let's examine the residuals of the logistic regression model to check for non-linearities.

```{r residual_analysis}
plot_residuals <- function(model, dataTrain, var_name) {
  # Calculate residuals on the full training set
  # We use the downsampled model to predict on the original training data
  probs_train <- predict(model, dataTrain, type = "prob")[, "Yes"]
  y_train_num <- ifelse(dataTrain$Diabetes_binary == "Yes", 1, 0)

  # Calculate Pearson residuals
  # Avoid division by zero if probs are 0 or 1 (add small epsilon)
  epsilon <- 1e-6
  probs_train <- pmax(pmin(probs_train, 1 - epsilon), epsilon)
  residuals_pearson <- (y_train_num - probs_train) / sqrt(probs_train * (1 - probs_train))

  # gut check: residuals should be close to 0 for linear relationships
  cat("Gut Check: Residuals should be close to 0 for linear relationships\n")
  cat("Mean Residual: ", mean(residuals_pearson), "\n")

  # Plot Residuals vs BMI
  # We use a sample of points to avoid overplotting if the dataset is huge
  set.seed(123)
  sample_idx <- sample(1:nrow(dataTrain), min(5000, nrow(dataTrain)))

  plot(dataTrain[[var_name]][sample_idx], residuals_pearson[sample_idx],
    main = paste("Pearson Residuals vs", var_name),
    xlab = var_name, ylab = "Pearson Residuals",
    pch = 20, col = rgb(0, 0, 0, 0.2)
  )
  lines(lowess(dataTrain[[var_name]][sample_idx], residuals_pearson[sample_idx]), col = "red", lwd = 2)
}

plot_deviance_residuals <- function(model, dataTrain, var_name) {
  # Predict probabilities on the full training set
  probs_train <- predict(model, dataTrain, type = "prob")[, "Yes"]
  y_train_num <- ifelse(dataTrain$Diabetes_binary == "Yes", 1, 0)

  # Avoid probabilities exactly 0 or 1
  epsilon <- 1e-6
  probs_train <- pmax(pmin(probs_train, 1 - epsilon), epsilon)

  # Deviance residuals for Bernoulli / binomial
  # r_i = sign(y_i - p_i) * sqrt(2 * [ y_i log(y_i/p_i) + (1-y_i) log((1-y_i)/(1-p_i)) ])
  # Handle y_i = 0 or 1 explicitly so we never take log(0)
  term1 <- ifelse(y_train_num == 0,
    0,
    y_train_num * log(y_train_num / probs_train)
  )
  term2 <- ifelse(y_train_num == 1,
    0,
    (1 - y_train_num) * log((1 - y_train_num) / (1 - probs_train))
  )

  dev_i <- 2 * (term1 + term2)
  residuals_deviance <- sign(y_train_num - probs_train) * sqrt(dev_i)

  # Gut check
  cat("Gut Check: Deviance residuals should be roughly symmetric around 0\n")
  cat("Mean Residual: ", mean(residuals_deviance), "\n")

  # Plot Residuals vs chosen variable
  set.seed(123)
  sample_idx <- sample(1:nrow(dataTrain), min(5000, nrow(dataTrain)))

  plot(dataTrain[[var_name]][sample_idx], residuals_deviance[sample_idx],
    main = paste("Deviance Residuals vs", var_name),
    xlab = var_name, ylab = "Deviance Residuals",
    pch = 20, col = rgb(0, 0, 0, 0.2)
  )
  lines(lowess(dataTrain[[var_name]][sample_idx], residuals_deviance[sample_idx]),
    col = "red", lwd = 2
  )
}
```

Check BMI, MentHlth, GenHlth, PhysHlth, Age, Education, Income

```{r}
plot_deviance_residuals(model_glm, dataTrain, "BMI")
```

# Results and Evaluation

We evaluate the models on the holdout test set using ROC-AUC, Sensitivity (Recall), and Specificity.

```{r evaluation}
# Function to calculate metrics
calc_metrics <- function(model, test_data) {
  probs <- predict(model, test_data, type = "prob")[, "Yes"]
  preds <- predict(model, test_data, type = "raw")

  cm <- confusionMatrix(preds, test_data$Diabetes_binary, positive = "Yes")
  roc_obj <- roc(test_data$Diabetes_binary, probs)
  auc_val <- auc(roc_obj)

  list(
    CM = cm,
    AUC = auc_val,
    F1 = cm$byClass["F1"],
    Recall = cm$byClass["Sensitivity"],
    Precision = cm$byClass["Precision"]
  )
}

metrics_glm <- calc_metrics(model_glm, dataTest)
metrics_down <- calc_metrics(model_glm_down, dataTest)
metrics_up <- calc_metrics(model_glm_up, dataTest)
metrics_lasso <- calc_metrics(model_lasso, dataTest)
metrics_lasso_1se <- calc_metrics(model_lasso_1se, dataTest)

# Display Comparison
results_df <- data.frame(
  Model = c("GLM (Standard)", "GLM (Downsample)", "GLM (Upsample)", "LASSO", "LASSO (1SE + Simplified)"),
  AUC = c(metrics_glm$AUC, metrics_down$AUC, metrics_up$AUC, metrics_lasso$AUC, metrics_lasso_1se$AUC),
  Recall = c(metrics_glm$Recall, metrics_down$Recall, metrics_up$Recall, metrics_lasso$Recall, metrics_lasso_1se$Recall),
  F1 = c(metrics_glm$F1, metrics_down$F1, metrics_up$F1, metrics_lasso$F1, metrics_lasso_1se$F1)
)

knitr::kable(results_df, digits = 3, caption = "Model Performance Comparison")

# Display Confusion Matrices for each model
metrics_glm$CM
metrics_down$CM
metrics_up$CM
metrics_lasso$CM
metrics_lasso_1se$CM
```

### Feature Importance

We analyze the multiplicative impact of features on the odds of diabetes using the coefficients from each of the models.

```{r feature_importance, fig.height=6, fig.width=8}
# Extract coefficients
plot_importances <- function(model, model_name) {
  coefs <- coef(model$finalModel)
  odds_ratios <- exp(coefs)
  odds_df <- data.frame(Feature = names(odds_ratios), OddsRatio = odds_ratios) %>%
    filter(Feature != "(Intercept)") %>%
    arrange(desc(OddsRatio))

  ggplot(odds_df, aes(
    x = reorder(Feature, OddsRatio),
    y = OddsRatio,
    fill = OddsRatio > 1
  )) +
    geom_col() +
    scale_fill_manual(values = c(`TRUE` = "steelblue", `FALSE` = "red")) +
    coord_flip() +
    labs(
      title = paste0("Feature Importance (Odds Ratios) - ", model_name),
      x = "Feature",
      y = "Multiplicative Impact on Odds",
      fill = "Odds > 1"
    ) +
    theme_minimal()
}

plot_importances(model_glm, "GLM (Standard)")
```

```{r}
plot_importances(model_lasso_1se, "GLM (Lasso + 1SE)")
```

```{r}
plot_importances(model_glm_down, "GLM (Downsample)")
```

# Fairness Analysis

We examine the model's performance across different demographic groups to ensure fairness. We will use the Downsampled GLM for this analysis.

```{r fairness_analysis}
# Define demographic variables to check
demo_vars <- c("Sex", "Age", "Education", "Income")

# Function to print confusion matrix for subgroups
analyze_fairness <- function(var_name, model, data) {
  cat(paste0("\n### Fairness Analysis by ", var_name, "\n"))
  groups <- unique(data[[var_name]])

  for (g in sort(groups)) {
    sub_data <- data[data[[var_name]] == g, ]
    if (nrow(sub_data) > 0) {
      preds <- predict(model, sub_data)
      cm <- confusionMatrix(preds, sub_data$Diabetes_binary, positive = "Yes")
      cat(paste0("\n#### Group: ", g, "\n"))
      print(cm$table)
      cat(paste0("Sensitivity: ", round(cm$byClass["Sensitivity"], 3), "\n"))
      cat(paste0("Specificity: ", round(cm$byClass["Specificity"], 3), "\n"))
      cat(paste0("Balanced Accuracy: ", round(cm$byClass["Balanced Accuracy"], 3), "\n"))
    }
  }
}

# Run fairness analysis
for (var in demo_vars) {
  if (var %in% names(dataTest)) {
    analyze_fairness(var, model_glm_down, dataTest)
  }
}
```

```{r}
for (var in demo_vars) {
  if (var %in% names(dataTest)) {
    analyze_fairness(var, model_glm, dataTest)
  }
}
```

```{r}
for (var in demo_vars) {
  if (var %in% names(dataTest)) {
    analyze_fairness(var, model_lasso_1se, dataTest)
  }
}
```

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
